Competing Bandits

### 2019-08-14 journal paper/ec19paper/content/introduction.tex
the latex compiles

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with a first-mover. Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``incumbent" (the first-mover in temporary monopoly) to permanent duopoly to ``entrant" (late-arriver in temporary monopoly).%
\footnote{\asedit{We consider the ``permanent monopoly" scenario for comparison only, without presenting any findings. We just assume that a monopolist chooses
the greedy algorithm, because it is easier to deploy in practice. Implicitly, users have no ``outside option": the service provided is an improvement over not having it (and therefore the monopolist is not incentivized to deploy better learning algorithms). This is plausible with free ad-supported platforms such as Yelp or Google.}}\footnote{\gaedit{The most common measures of market ``competitiveness" such as the Lerner Index or the Herfindahl-Hirschman Index of a market rely on ex-post observable attributes of a market such as prices or market shares \cite{tirole1988theory}. However, since we have no prices and market shares are an endogenous element of our model, we simply define ``competitiveness" in terms of the number of firms and timing of entry into the market.}}

@@

We also consider an alternative choice rule with explicit noise/randomness: a small fraction of users choose a firm uniformly at random.%
\footnote{\asedit{Reputation scores already introduce some noise into users' choices. However, the amount of noise due to this channel is typically small, both in our simulations and in practice, because reputation signals average over many datapoints.}} We confirm the theoretical intuition that better algorithms prevail if the \gaedit{expected number of ``random" users is sufficiently large.}\gadelete{parameter -- essentially, the expected number ``random users" --is sufficiently large.} However, we find that this effect is negligible for some smaller but still ``relevant" parameter values.

@@

\asedit{While amenable to simulations, our model appears difficult to analyze. This is for several reasons:
%
intricate feedback loop from performance to reputations to users to performance;
%
mean reputation, most connected to our intuition, is sometimes \gaedit{a} bad predictor in competition (see Sections~\ref{sec:isolation} and~\ref{sec:revisited});
%
mathematical tools from regret-minimization would only produce ``asymptotic" results, which do not seem to suffice. We believe that resolving first-order theoretical questions would not add much value to this paper, because of \cite{CompetingBandits-itcs18}.\gadelete{We are in \gaedit{the} realm of ``asymptotic" theorems about stylized economic models, which provide some mathematical intuition but no direct bearing on the world. But \cite{CompetingBandits-itcs18} already has an elaborate set of such results, with similar high-level conclusions. So, one would need to go beyond the first-order conclusions to add value.}}

@@

>>>>

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with a first-mover. Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``incumbent" (the first-mover in temporary monopoly) to permanent duopoly to ``entrant" (late-arriver in temporary monopoly).%
\footnote{\asedit{We consider the ``permanent monopoly" scenario for comparison only, without presenting any findings. We just assume that a monopolist chooses
the greedy algorithm, because it is easier to deploy in practice. Implicitly, users have no ``outside option": the service provided is an improvement over not having it (and therefore the monopolist is not incentivized to deploy better learning algorithms). This is plausible with free ad-supported platforms such as Yelp or Google.}}

@@

We also consider an alternative choice rule with explicit noise/randomness: a small fraction of users choose a firm uniformly at random.%
\footnote{\asedit{Reputation scores already introduce some noise into users' choices. However, the amount of noise due to this channel is typically small, both in our simulations and in practice, because reputation signals average over many datapoints.}} We confirm the theoretical intuition that better algorithms prevail if the expected number of ``random" users is sufficiently large. However, we find that this effect is negligible for some smaller but still ``relevant" parameter values.

@@

While amenable to simulations, our model appears difficult to analyze. This is for several reasons:
%
intricate feedback loop from performance to reputations to users to performance;
%
mean reputation, most connected to our intuition, is sometimes a bad predictor in competition (see Sections~\ref{sec:isolation} and~\ref{sec:revisited});
%
mathematical tools from regret-minimization would only produce ``asymptotic" results, which do not seem to suffice. Further, we believe that resolving first-order theoretical questions about our model would not add much value to this paper. \asedit{Indeed, we are in the realm of stylized economic models that provide mathematical intuition about the world, and \cite{CompetingBandits-itcs18} already has an elaborate analysis with similar high-level conclusions.}

@@

\asedit{The most common measures of market ``competitiveness" such as the Lerner Index or the Herfindahl-Hirschman Index of a market rely on ex-post observable attributes of a market such as prices or market shares \cite{tirole1988theory}. However, neither is applicable to our setting: in our model, there are no prices, and market shares are endogenous.}

### 2019-08-14 journal paper/ec19paper/content/inverted_u.tex
the latex compiles

\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
   & $\TS$  & $\DEG$  & $\DG$ \\ \hline
$\TS$
    & \makecell{\textbf{0.003}$\pm$0.003}
    & \makecell{\textbf{0.083}$\pm$0.02}
    & \makecell{\textbf{0.17}$\pm$0.02} \\\hline
$\DEG$
    & \makecell{\textbf{0.045}$\pm$0.01}
    & \makecell{\textbf{0.25}$\pm$0.02}
    & \makecell{\textbf{0.23}$\pm$0.02} \\\hline
$\DG$
    & \makecell{\textbf{0.12}$\pm$0.02}
    & \makecell{\textbf{0.36}$\pm$0.03}
    & \makecell{\textbf{0.3}$\pm$0.02} \\\hline
\end{tabular}
\caption{{\bf Temporary monopoly}, with $X=200$ (and $T_0=20$), for the Heavy-Tail MAB instance. Each cell describes the duopoly game between the entrant's algorithm (the row) and the incumbent's algorithm (the column). The cell specifies the entrant's market share (fraction of rounds in which it was chosen) for the rounds in which he was present. We give the average (in bold) and the 95\% confidence interval. NB: smaller average is better for the incumbent.}
\label{tab:ht-incum}
\end{table}
\OMIT{
\begin{table}[h]
\centering
\caption{Heavy Tail}
\begin{tabular}{|c|c|c|c|}
\hline
   & $\TS$  & $\DEG$  & $\DG$ \\ \hline
$\TS$
    & \makecell{\textbf{0.50, 0.50}}
    & \makecell{\textbf{0.3, 0.7}}
    & \makecell{\textbf{0.29, 0.71}} \\\hline
$\DEG$
    & \makecell{\textbf{0.7, 0.3}}
    & \makecell{\textbf{0.50, 0.50}}
    & \makecell{\textbf{0.38, 0.62}} \\\hline
$\DG$
    & \makecell{\textbf{0.71, 0.29}}
    & \makecell{\textbf{0.62, 0.38}}
    & \makecell{\textbf{0.50, 0.50}} \\\hline
\end{tabular}
\end{table}

@@

\begin{figure}
\includegraphics[scale=0.3]{figures/unif_eq_welfare}
\includegraphics[scale=0.3]{figures/ht_eq_welfare}
\caption{Smoothed welfare plots resulting from equilibrium strategies in the different market structures. Note that welfare at $t = 0$ incorporates the regret incurred during the incumbent and warm start periods.}
\label{eq_regret}
\end{figure}

@@

Finding~\ref{find:welfare}(b) is interesting because, in equilibrium, both firms play \DG in both settings, and one might conjecture that the welfare should increase with the number of firms playing \DG. Indeed, one run of \DG may get stuck on a bad arm. However, two firms independently playing \DG are less likely to get stuck simultaneously. If one firm gets stuck and the other does not, then the latter should attract most agents, leading to improved welfare.
\begin{figure}
\includegraphics[scale=0.3]{figures/eeog_vs_num_ht}
\includegraphics[scale=0.3]{figures/ht_many_firm_welfare}\\
\caption{Average welfare and \Eeog as we increase \#firms playing \DG}
\label{many_firm_welfare}
\end{figure}

>>>>

\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
   & $\TS$  & $\DEG$  & $\DG$ \\ \hline
$\TS$
    & \makecell{\textbf{0.003}$\pm$0.003}
    & \makecell{\textbf{0.083}$\pm$0.02}
    & \makecell{\textbf{0.17}$\pm$0.02} \\\hline
$\DEG$
    & \makecell{\textbf{0.045}$\pm$0.01}
    & \makecell{\textbf{0.25}$\pm$0.02}
    & \makecell{\textbf{0.23}$\pm$0.02} \\\hline
$\DG$
    & \makecell{\textbf{0.12}$\pm$0.02}
    & \makecell{\textbf{0.36}$\pm$0.03}
    & \makecell{\textbf{0.3}$\pm$0.02} \\\hline
\end{tabular}
\captionsetup{labelformat=empty}
\caption{User share of row player (entrant), 200 round head-start, Heavy-Tail Instance}
%\caption{{\bf Temporary monopoly}, with $X=200$ (and $T_0=20$), for the Heavy-Tail MAB instance. Each cell describes the duopoly game between the entrant's algorithm (the row) and the incumbent's algorithm (the column). The cell specifies the entrant's market share (fraction of rounds in which it was chosen) for the rounds in which he was present. We give the average (in bold) and the 95\% confidence interval. NB: smaller average is better for the incumbent.}
\label{tab:ht-incum}
\end{table}
\OMIT{
\begin{table}[h]
\centering
\caption{Heavy Tail}
\begin{tabular}{|c|c|c|c|}
\hline
   & $\TS$  & $\DEG$  & $\DG$ \\ \hline
$\TS$
    & \makecell{\textbf{0.50, 0.50}}
    & \makecell{\textbf{0.3, 0.7}}
    & \makecell{\textbf{0.29, 0.71}} \\\hline
$\DEG$
    & \makecell{\textbf{0.7, 0.3}}
    & \makecell{\textbf{0.50, 0.50}}
    & \makecell{\textbf{0.38, 0.62}} \\\hline
$\DG$
    & \makecell{\textbf{0.71, 0.29}}
    & \makecell{\textbf{0.62, 0.38}}
    & \makecell{\textbf{0.50, 0.50}} \\\hline
\end{tabular}
\end{table}

@@

\begin{figure}
\includegraphics[scale=0.3]{ec19paper/figures/unif_eq_welfare}
\includegraphics[scale=0.3]{ec19paper/figures/ht_eq_welfare}
\caption{Smoothed welfare plots resulting from equilibrium strategies in the different market structures. Note that welfare at $t = 0$ incorporates the regret incurred during the incumbent and warm start periods.}
\label{eq_regret}
\end{figure}

@@

Finding~\ref{find:welfare}(b) is interesting because, in equilibrium, both firms play \DG in both settings, and one might conjecture that the welfare should increase with the number of firms playing \DG. Indeed, one run of \DG may get stuck on a bad arm. However, two firms independently playing \DG are less likely to get stuck simultaneously. If one firm gets stuck and the other does not, then the latter should attract most agents, leading to improved welfare.
\begin{figure}
\includegraphics[scale=0.3]{ec19paper/figures/eeog_vs_num_ht}
\includegraphics[scale=0.3]{ec19paper/figures/ht_many_firm_welfare}\\
\caption{Average welfare and \Eeog as we increase \#firms playing \DG}
\label{many_firm_welfare}
\end{figure}

### 2019-08-14 journal paper/ec19paper/content/perf_in_iso.tex
the latex compiles

\begin{figure}
\includegraphics[scale=0.35]{figures/nih_iso_mean}
%\caption*{\tiny{Mean reputation trajectory The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of a given algorithm on a given instance and then average this value across all the runs. The shaded area display 95\% confidence intervals.}}
\caption{Mean reputation trajectories for Needle-in-Haystack. The shaded area shows 95\% confidence intervals.}
\label{prelim_means}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[scale=0.35]{figures/relative_uniform_annotated_plot}
\includegraphics[scale=0.35]{figures/relative_nih_ts_dg.png}
%\caption*{\tiny{The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of both of the algorithms on a given instance and then calculate the proportion of runs where $\TS \geq \DG$. The shaded area display 95\% confidence intervals.}}
\caption{Relative reputation trajectory for $\TS$ vs $\DG$, on Uniform instance (top) and Needle-in-Haystack instance (bottom). Shaded area display 95\% confidence intervals.}
\label{relative_rep_plots}
\end{figure}

>>>>

\begin{figure}
\includegraphics[scale=0.35]{ec19paper/figures/nih_iso_mean}
%\caption*{\tiny{Mean reputation trajectory The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of a given algorithm on a given instance and then average this value across all the runs. The shaded area display 95\% confidence intervals.}}
\caption{Mean reputation trajectories for Needle-in-Haystack. The shaded area shows 95\% confidence intervals.}
\label{prelim_means}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[scale=0.35]{ec19paper/figures/relative_uniform_annotated_plot}
\includegraphics[scale=0.35]{ec19paper/figures/relative_nih_ts_dg.png}
%\caption*{\tiny{The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of both of the algorithms on a given instance and then calculate the proportion of runs where $\TS \geq \DG$. The shaded area display 95\% confidence intervals.}}
\caption{Relative reputation trajectory for $\TS$ vs $\DG$, on Uniform instance (top) and Needle-in-Haystack instance (bottom). Shaded area display 95\% confidence intervals.}
\label{relative_rep_plots}
\end{figure}

### 2019-08-14 journal paper/journal_main.tex
the latex compiles

\usepackage{acm-ec-19}
\usepackage{tikz} % SW: Added for plots
\usepackage{subfiles}
\usepackage[toc,page]{appendix}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{natbib} % cannot use with AAAI! A.S.

@@

% Alex' macro
\usepackage{xspace}
\newcommand{\OMIT}[1]{}
\newcommand{\xhdr}[1]{\vspace{1mm} \noindent{\bf #1}}
\newcommand{\ie}{{\em i.e.,~\xspace}}
\newcommand{\eg}{{\em e.g.,~\xspace}}

\newcommand{\term}[1]{\ensuremath{\mathtt{#1}}\xspace}
\newcommand{\TS}{\term{TS}}
\newcommand{\DEG}{\term{DEG}}
\newcommand{\DG}{\term{DG}}
\newcommand{\AD}{\term{AD}}
\newcommand{\EGR}{\term{EGR}}
\newcommand{\GR}{\term{GR}}
\newcommand{\Beta}{\term{Beta}} % for Beta distribution
\newcommand{\Eeog}{\term{EoG}} % shorthand for "effective end of game"

\newcommand{\eps}{\varepsilon}

@@

\newcommand{\HMR}{\term{HMR}}
\newcommand{\HM}{\term{HM}}
% a very useful package for edits and comments, from David Kempe (USC)
\usepackage{color-edits}
%\usepackage[suppress]{color-edits}  % use this to suppress the package
\addauthor{as}{red}    % as for Alex
\addauthor{ga}{blue}  % ga for Guy
\addauthor{sw}{orange} % sw for Steven
% e.g. for Alex, provides \asedit{}, \ascomment{} and \asdelete{}.
\setlength{\tabcolsep}{8pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5} % Default value: 1

@@

\begin{document}
\title[The Perils of Exploration under Competition: A Computational Modeling Approach]
{The Perils of Exploration under Competition: \\ A Computational Modeling Approach}

 \author{Guy Aridor}
 \affiliation{ \institution{Columbia University}}
 \author{Kevin Liu}
 \affiliation{\institution{Columbia University}}
 \author{Aleksandrs Slivkins}
 \affiliation{\institution{Microsoft Research, New York City}}
 \author{Zhiwei Steven Wu}
 \affiliation{\institution{University of Minnesota - Twin Cities}}

 \authorsaddresses{%
  Authors' addresses: Guy Aridor, Columbia University, New York, NY, g.aridor@columbia.edu; Kevin Liu, Columbia University, New York, NY, kevin.liu@columbia.edu; Aleksandrs Slivkins, Microsoft Research, New York, NY, slivkins@microsoft.com; Zhiwei Steven Wu, University of Minnesota - Twin Cities, Minneapolis, MN, zsw@umn.edu}

@@

\subfile{content/introduction}

\subfile{content/model}

\subfile{content/perf_in_iso}

\subfile{content/inverted_u}

\subfile{content/barriers}

\subfile{content/revisited}

\subfile{content/non_greedy_choice}

\subfile{content/conclusion}
\newpage
\bibliographystyle{acm}
\bibliography{refs,bib-abbrv-short,bib-bandits,bib-AGT,bib-slivkins}

\subfile{content/appendix_for_one_version}

>>>>

\usepackage{ec19paper/acm-ec-19}
\usepackage{itcs18paper/slivkins-setup,itcs18paper/slivkins-theorems}
\usepackage{tikz} % SW: Added for plots
\usepackage{subfiles}
\usepackage[toc,page]{appendix}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{natbib} % cannot use with AAAI! A.S.

@@

% Commands from EC 19 paper
\newcommand{\term}[1]{\ensuremath{\mathtt{#1}}\xspace}
\newcommand{\TS}{\term{TS}}
\newcommand{\DEG}{\term{DEG}}
\newcommand{\DG}{\term{DG}}
\newcommand{\AD}{\term{AD}}
\newcommand{\EGR}{\term{EGR}}
\newcommand{\GR}{\term{GR}}
\newcommand{\Beta}{\term{Beta}} % for Beta distribution
\newcommand{\Eeog}{\term{EoG}} % shorthand for "effective end of game"

@@

\newcommand{\HMR}{\term{HMR}}
\newcommand{\HM}{\term{HM}}

% commands from ITCS 18 paper

\DeclareMathOperator*{\Expectation}{\mathbb{E}}
\newcommand{\Ex}[2]{\Expectation_{#1}\left[#2\right]}

% notation
%%% advanced notation
\newcommand{\OPT}{\term{OPT}}
\newcommand{\rew}{\term{rew}}  % Bayesian-expected reward after n local rounds
\newcommand{\PMR}{\term{PMR}} % posterior mean reward
\newcommand{\support}{\term{support}}

\newcommand{\BIR}{\term{BIR}} % Bayesian Instantaneous Regret
\newcommand{\regret}{R^{\term{inst}}} % regret
\newcommand{\regretWC}{\regret_{\term{wc}}} % regret

% response function
\newcommand{\respF}{f_{\term{resp}}}
\newcommand{\respEps}{\eps_\term{resp}}

\newcommand{\BReg}{\term{BReg}}
\newcommand{\HardMax}{\term{HardMax}}
\newcommand{\HardMaxRandom}{\term{HardMax\&Random}}
\newcommand{\SoftMaxRandom}{\term{SoftMax}}
\newcommand{\Uniform}{\term{Uniform}}
\newcommand{\Random}{\term{Random}}

\newcommand{\StaticGreedy}{\term{StaticGreedy}}
\newcommand{\DynGreedy}{\term{DynamicGreedy}}
\newcommand{\DynamicGreedy}{\term{DynamicGreedy}}

\newcommand{\termSub}[2]{\ensuremath{\mathtt{#1}_{#2}}\xspace}
\newcommand{\alg}[1][]{\termSub{alg}{#1}}
%\newcommand{\prin}[1][]{\termSub{prin}{#1}}  % principal
\newcommand{\agent}[1][]{\termSub{agent}{#1}}

% priors and posteriors
\newcommand{\prior}{\ensuremath{\mP}\xspace}
\newcommand{\priorMu}{\ensuremath{\prior_\mathtt{mean}}\xspace}
\newcommand{\posteriorN}[2]{\mN_{#1,#2}}  % \posteriorN{principal}{round}

% rationality / innovation / competition
\newcommand{\termTXT}[1]{{\em {#1}}\xspace}

\newcommand{\rationality}{\termTXT{rationality}}
\newcommand{\Rationality}{\termTXT{Rationality}}
\newcommand{\innovation}{\termTXT{innovation}}
\newcommand{\Innovation}{\termTXT{Innovation}}
\newcommand{\competition}{\termTXT{competition}}
\newcommand{\Competition}{\termTXT{Competition}}
\newcommand{\competitiveness}{\termTXT{competitiveness}}
\newcommand{\exploration}{\termTXT{exploration}}
\newcommand{\Exploration}{\termTXT{Exploration}}
% a very useful package for edits and comments, from David Kempe (USC)
\usepackage{ec19paper/color-edits}
%\usepackage[suppress]{color-edits}  % use this to suppress the package
\addauthor{as}{red}    % as for Alex
\addauthor{ga}{blue}  % ga for Guy
\addauthor{sw}{orange} % sw for Steven
% e.g. for Alex, provides \asedit{}, \ascomment{} and \asdelete{}.
\setlength{\tabcolsep}{8pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5} % Default value: 1

@@

\begin{document}
\title[Competing Bandits]
{Competing Bandits}

 \author{Guy Aridor}
 \affiliation{ \institution{Columbia University}}
 \author{Aleksandrs Slivkins}
 \affiliation{\institution{Microsoft Research, New York City}}
 \author{Zhiwei Steven Wu}
 \affiliation{\institution{University of Minnesota - Twin Cities}}

 \authorsaddresses{%
  Authors' addresses: Guy Aridor, Columbia University, New York, NY, g.aridor@columbia.edu; Aleksandrs Slivkins, Microsoft Research, New York, NY, slivkins@microsoft.com; Zhiwei Steven Wu, University of Minnesota - Twin Cities, Minneapolis, MN, zsw@umn.edu}

@@

\section{Introduction}
\label{sec:intro}
\subfile{itcs18paper/sec-intro}

\section{Related work}
\label{sec:related-work}
\subfile{itcs18paper/sec-related-work}

\section{Our model and preliminaries}
\label{sec:model}
\subfile{itcs18paper/sec-model}

\section{Full rationality (HardMax)}
\label{sec:rational}
\subfile{itcs18paper/sec-rational}

\section{Relaxed rationality: HardMax \& Random}
\label{sec:random}
\subfile{itcs18paper/sec-random}

\section{SoftMax response function}
\label{sec:soft}
\subfile{itcs18paper/sec-soft}

\section{Economic implications}
\label{sec:welfare}
\subfile{itcs18paper/sec-welfare}

\subfile{ec19paper/content/introduction}

\subfile{ec19paper/content/model}

\subfile{ec19paper/content/perf_in_iso}

\subfile{ec19paper/content/inverted_u}

\subfile{ec19paper/content/barriers}

\subfile{ec19paper/content/revisited}

\subfile{ec19paper/content/non_greedy_choice}

\subfile{ec19paper/content/conclusion}

\newpage
\bibliographystyle{acm}
\bibliography{bib-ML,refs,bib-abbrv-short,bib-bandits,bib-AGT,bib-slivkins}

\subfile{ec19paper/content/appendix_for_one_version}
